Space is the Place: Effects of Continuous Spatial Structure on Analysis of Population Genetic Data

### 2020-02-10 ms/second-review-responses.tex

\reply{
Another important point set of points on relating neighborhood sizes from different models here. In our simulation the average x axis displacement of a locus along any link in the pedigree is $\sigma$ for the "maternal" material (i.e. deriving from the parent from which dispersal is calculated), but for the "paternal" parent it is the sum of two independent normal draws each with variance $\sigma^2$. Averaging over a random series of maternal and paternal transmissions (because our simulated genome is autosomal), variance in axial position across pedigree links is then $(3/2)\sigma^2$ and the expected displacement $\sqrt{3/2}\sigma$. So from the perspective of genes sampled at the present time our neighborhood size calculations are lower than those in models like those described in \cite{Rousset1997} by a factor of (3/2) because of differences in the breeding model.

This inflation in neighborhood size is then partially offset by our use of census rather than "effective" population size when calculating population density, because $N_e$ of our spatial model estimated from $\pi$ in random-mating spatial models is roughly (2/3) the census size. We chose this method for density and sigma calculations because we sought to parameterize our model with values that could in theory be observed in real populations. We also find the concept of "effective population size" in this context even more confusing than usual -- as we show, spatial processes in both evolution and sampling radically shift the distribution of coalescence times and the variance in offspring numbers in ways that mimic the effects of classic $N_{e}$ scaling in random-mating models, but are driven by fundamentally different processes.

Thus relative to models like \cite{Rousset1997} our definition of $\sigma$ is probably too low and $\rho$ probably too high; while relative to neighborhood sizes calculated from mark-recapture studies measuring parent-offspring distances our figures should be consistent. We have expanded the caveats at the head of the discussion paragraph interpreting empirical studies noting that definitions of $\rho$ in ecological vs genetic studies mean results should be interpreted with caution \llname{rhocaveat}; however, as we mainly interpret these figures in terms of order-of-magnitude differences (roughly less than 100, 100-1000, and over 1000) we think our results are general enough to be robust to these modeling differences while still giving a useful yardstick. Indeed for many of the empirical studies the range of NS values given is larger than the differences we expect from the parameter definitions described above. We also think that our study of lattice simulations in appendix 1 demonstrates nicely how qualitatively similar simulations can lead to quite different patterns of genetic diversity at a given neighborhood size as a result of differences in the underlying breeding model.

### 2020-02-11 ms/second-review-responses.tex

\pagebreak
\setcounter{page}{1}

%%%%%%%%%%%%%%
\reviewersection{AE}

The point of R3 on the bimodality in the realized dispersal distance in your model is interesting. You can make this more explicit in the paper and describe how there are realistic biological scenarios that motivate such a model (many/most plants) and demarcate how well you think you can generalize to others (the new appendix already helps in that regard). It strikes me your model has relevance for patrilocal/matrilocal systems where the paternal/maternal parent-offspring distances differ - though your model is hermaphroditic so it's an imperfect analogy. Perhaps, to explore more typical animal mating patterns, you can quickly check the impact for one or two outcomes of changing the simulation such that offspring are dispersed around the midpoint of the parental origins, as this would remove the bimodality. That said, please be sure we respect the work done is extensive and don't mean to invite/require a complete re-working. The core tension is between ensuring your results are generalizable if that is what you wish to claim; versus specific to a particular model that is biologically motivated. It's particularly key for the material on GWAS, where you are explicitly trying to address human GWAS results: those sections will be read and taken at face value by a large community even if this simulation scheme introduces some discrepancies from what might more realistically emerge. So at a minimum, some caveats directly in the GWAS discussion section would help.

\reply{Thank you for your comments. We have responded to each reviewer below. In response to reviewer 2 we have added several caveats to the GWAS section noting other sources of error such as allele frequency and LD variation among populations. Reviewer 3 raised two good points on our definition of neighborhood size and the interpetation of $\sigma$ -- first, that $\sigma$ refers to one- rather than two-dimensional distances between parent and offspring (this is consistent with previous theoretical work, and we have corrected our definition in the text), and second, that the use of separate draws for mate selection and dispersal means the total distance traveled by a locus along a pedigree link is greater for "paternal" than "maternal" material (taking "maternal" to mean the parent from which dispersal is calculated). Because our simulated genome is autosomal and individuals hermaphroditic and non-selfing, a locus will experience a mix of "maternal" and "paternal" transmission events along its genealogy. So we see this as a scaling issue rather than one of a "bimodal" dispersal kernal per se -- the expected axial displacement of a locus along a pedigree link is $\sqrt{3/2} \sigma$, rather than $\sigma$. However we note that in terms of summary statistics the effects of scaling $\sigma$ by $sqrt{3/2}$ are weaker than those generated by factors like changing landscape resolution in lattice vs. continuous models (e.g. Appendix 1), so we don't think this significantly changes our interpretation of model outputs. }

@@

\begin{point}{Line 686. }
 I don't know that the results of Berg and Sohail et al. are indicating a fundamental issue with PC correction, rather than a problem with meta-analysis.
\end{point}

\reply{
We think Berg et al. shows that population structure leaves signals in GWAS that resemble those of polygenic selection, and given that this was most apparent in GIANT data employing PCA corrections, indicates a problem with the GWAS that make up the GIANT study at least as much as any issues with meta-analysis. 
}

\begin{point}{Line 689-92. }
 Is the result of Martin et al. really due to population stratification? There seem to be a number of factors affecting transferability of PRS (LD differences, effect size heterogeneity, differences in causal variants etc..), but I don't know that anyone thinks residual structure is the major one, particularly for eg. UKB summary stats.
\end{point}

\reply{This is a good point -- differences in allele frequency and LD structure certainly have significant effects on PRS transferability (at least when the GWAS is based on SNP assays rather than whole-genome sequencing). We think the population structure issue is important to point to here because it will not be solved by simply generating more detailed genomic data from existing samples, but have added a caveat noting these other important potential sources of error at the head of that sentence. \llname{ldcaveats}
}

@@

It is important that the authors clarify this last point because most results presented in this study rely on Wright's neighborhood size, which derives from a model where dispersal of individuals and chromosomes follow well-defined spatial dynamics. A sensible comparison of neighborhood sizes requires that dispersal (along with effective population density) is quantified in a coherent way across different models.

Additionally, the neighborhood size as defined by the authors (see line 211) is a function of the census density rather than the effective density of individuals. I am therefore not fully convinced that the range of values for this parameter is comparable to that given in Table 1.
\end{point}

\reply{
Another important point set of points on relating neighborhood sizes from different models here. In our simulation the average x axis displacement of a locus along any link in the pedigree is $\sigma$ for the "maternal" material (i.e. deriving from the parent from which dispersal is calculated), but for the "paternal" parent it is the sum of two independent normal draws each with variance $\sigma^2$. Averaging over a random series of maternal and paternal transmissions (because our simulated genome is autosomal), variance in axial position across pedigree links is then $(3/2)\sigma^2$ and the expected displacement $\sqrt{3/2}\sigma$. So from the perspective of genes sampled at the present time our neighborhood size calculations are lower than those in models like those described in \cite{Rousset1997} by a factor of (3/2) because of differences in the breeding model.

This inflation in neighborhood size is then partially offset by our use of census rather than "effective" population size when calculating population density, because $N_e$ of our spatial model estimated from $\pi$ in random-mating spatial models is roughly 2/3 the census size. We chose this method for density and sigma calculations because we sought to parameterize our model with values that could be observed in real populations. We also find the concept of "effective population size" in this context even more confusing than usual -- as we show, spatial processes in both evolution and sampling radically shift the distribution of coalescence times and the variance in offspring numbers in ways that mimic the effects of classic $N_{e}$ scaling in random-mating models, but are driven by fundamentally different processes.

Thus relative to models like \cite{Rousset1997} our definition of $\sigma$ is probably too low and $\rho$ probably too high such that the neighborhood sizes we show in plots are roughly consistent with previous theoretical work; while relative to neighborhood sizes calculated from mark-recapture studies measuring parent-offspring distances our figures should be consistent to the extent that the breeding and dispersal system matches our simulation. We have expanded the caveats at the head of the discussion paragraph interpreting empirical studies noting that definitions of $\rho$ in ecological vs genetic studies mean results should be interpreted with caution \llname{rhocaveat}; however, as we mainly interpret these figures in terms of order-of-magnitude differences (roughly less than 100, 100-1000, and over 1000) we think our results are general enough to be robust to these modeling differences while still giving a useful yardstick. Indeed for many of the empirical studies the range of NS values given is larger than the differences we expect from the parameter definitions described above. We also think that our study of lattice simulations in appendix 1 demonstrates nicely how qualitatively similar simulations can lead to quite different patterns of genetic diversity at a given neighborhood size as a result of differences in the underlying breeding model.

### 2020-02-11 ms/second-review-responses.tex

{\bf 
C.~J.~Battey, Peter Ralph, and Andrew Kern
}\\
\end{minipage}

\vskip 4em

\pagebreak
\setcounter{page}{1}

%%%%%%%%%%%%%%
\reviewersection{Decision Letter}

When you submit a revised manuscript please include a response to each of the reviewers' comments. It is most important that you address the following in a revised manuscript: 1) in response to reviewer 3, add text to help clarify how the reproduction scheme / dispersal model differs from models that do not induce the bimodality in the distance distribution the reviewer points out (see my comments below as well); 2) Respond to the concern about the definitions of sigma and neighborhood size; 3) Reviewer 2 has several minor comments that would be straightforward and helpful to address/add depth into your discussion (e.g., comments 2, 4, 5). 4) The other minor comments from each reviewer.

\reviewersection{AE Comments}
The point of R3 on the bimodality in the realized dispersal distance in your model is interesting. You can make this more explicit in the paper and describe how there are realistic biological scenarios that motivate such a model (many/most plants) and demarcate how well you think you can generalize to others (the new appendix already helps in that regard). It strikes me your model has relevance for patrilocal/matrilocal systems where the paternal/maternal parent-offspring distances differ - though your model is hermaphroditic so it's an imperfect analogy. Perhaps, to explore more typical animal mating patterns, you can quickly check the impact for one or two outcomes of changing the simulation such that offspring are dispersed around the midpoint of the parental origins, as this would remove the bimodality. That said, please be sure we respect the work done is extensive and don't mean to invite/require a complete re-working. The core tension is between ensuring your results are generalizable if that is what you wish to claim; versus specific to a particular model that is biologically motivated. It's particularly key for the material on GWAS, where you are explicitly trying to address human GWAS results: those sections will be read and taken at face value by a large community even if this simulation scheme introduces some discrepancies from what might more realistically emerge. So at a minimum, some caveats directly in the GWAS discussion section would help.

@@

\begin{point}{}
More importantly, I am still a bit confused about the reproduction scheme. My understanding is that, once mating takes place, the two parents involved (noted as P(1) and P(2)) each produce a Poisson number of offspring. The location of each offspring is determined by a Gaussian distribution of covariance matrix $\sigma^2 I$, centered on the location of the parent that ``produced' this offspring. Let $O(1,i)$ and $O(2,j)$ be the i-th and j-th offspring produced by $P(1)$ and $P(2)$ respectively. Now, if I am not mistaken, half of the genetic material of O(2,j) comes from P(1) (the other half coming from P(2)) while O(1,i) receives one chromosome from P(2) (and another one from its ``true' parent, i.e., P(1)). Hence, when it comes to the parent-offspring distance as measured from their genetic material, half of the chromosomes have a ``small' dispersal distance, equal to that of the parent-offspring distance, while the other half have a ``large' distance roughly equal to the parent-offspring distance plus the distance between mates.

It is important that the authors clarify this last point because most results presented in this study rely on Wright's neighborhood size, which derives from a model where dispersal of individuals and chromosomes follow well-defined spatial dynamics. A sensible comparison of neighborhood sizes requires that dispersal (along with effective population density) is quantified in a coherent way across different models.

Additionally, the neighborhood size as defined by the authors (see line 211) is a function of the census density rather than the effective density of individuals. I am therefore not fully convinced that the range of values for this parameter is comparable to that given in Table 1.
\end{point}

\reply{
These are an important point set of points on relating neighborhood sizes from different models which we have aimed to clarify. In our simulation the x axis displacement of a locus along any link in the pedigree is normally distributed with standard deviation $\sigma$ for the "maternal" material (i.e. deriving from the parent from which dispersal is calculated), but for the "paternal" parent it is the sum of two independent normal draws each with with standard deviation $\sigma$. Averaging over a random series of maternal and paternal transmissions (because our simulated genome is autosomal), variance in axial position across pedigree links is then $(3/2)\sigma^2$ and the expected displacement $\sqrt{3/2}\sigma$. So from the perspective of genes sampled at the present time our neighborhood size calculations are lower than those in models like those described in \cite{Rousset1997} by a factor of (3/2) because of differences in the breeding model.

@@

Thus relative to models like \cite{Rousset1997} our definition of $\sigma$ is probably too low and $\rho$ probably too high such that the neighborhood sizes we show in plots are roughly consistent with previous theoretical work; while relative to neighborhood sizes calculated from mark-recapture studies measuring parent-offspring distances our figures should be consistent to the extent that the breeding and dispersal system matches our simulation. We have expanded the caveats at the head of the discussion paragraph interpreting empirical studies noting that definitions of $\rho$ in ecological vs genetic studies mean results should be interpreted with caution \llname{rhocaveat}; however, as we mainly interpret these figures in terms of order-of-magnitude differences (roughly less than 100, 100-1000, and over 1000) we think our results are general enough to be robust to these modeling differences while still giving a useful yardstick. Indeed for many of the empirical studies the range of NS values given is larger than the differences we expect from the parameter definitions described above. We also think that our study of lattice simulations in appendix 1 demonstrates nicely how qualitatively similar simulations can lead to quite different patterns of genetic diversity at a given neighborhood size as a result of differences in the underlying breeding model.

\cjb{CJ: let's meet and talk again and talk about what edits to make here. I'm not sure what the best way to go is.}

}

\cjb{ADK: I made some small edits but I'm pretty happy about how it reads}

### 2020-02-11 ms/spaceness.tex
minor touch ups

In this work we simulated a single chromosome with size roughly comparable to one human chromosome. If we scale the number of false-positive associations identified in our analyses to a GWAS conducted on whole-genome data from humans, we would expect to see several thousand weak false-positive associations after PC corrections in a population with neighborhood sizes up to at least 1000 (which should include values appropriate for many human populations). Notably, very few of the spurious associations we identified would be significant at a conservative Bonferroni-adjusted $p$-value cutoff (see Figure \ref{fig:manhattan_plots}). This suggests that GWAS focused on finding strongly associated alleles for traits controlled by a limited number of variants in the genome are likely robust to the impacts of continuous spatial structure. However, methods that analyze the combined effects of thousands or millions of weakly associated variants such as polygenic risk scores \citep{Khera2018} are likely to be affected by subtle population structure. Indeed as recently identified in studies of genotype associations for human height in Europe \citep{Berg2018,Sohail2018}, PC regression GWAS in modern human populations do include residual signal of population structure in large-scale analyses of polygenic traits. In addition to error associated with varying patterns of linkage disequilibrium and allele frequency among populations, the confounding of environmental and genetic effects on phenotypes introduced by population structure is expected to lead to low predictive power when polygenic scores are generated for populations outside the original GWAS cohort, \llabel{ldcaveats}
%When attempting to make predictions across populations with different environmental exposures, polygenic risk scores affected by population structure can be expected to offer low predictive power, 
as was shown in a recent study finding lower polygenic score predictive power outside European populations \citep{Martin2019}.

>>>>

In this work we simulated a single chromosome with size roughly comparable to one human chromosome. If we scale the number of false-positive associations identified in our analyses to a GWAS conducted on whole-genome data from humans, we would expect to see several thousand weak false-positive associations after PC corrections in a population with neighborhood sizes up to at least 1000 (which should include values appropriate for many human populations). Notably, very few of the spurious associations we identified would be significant at a conservative Bonferroni-adjusted $p$-value cutoff (see Figure \ref{fig:manhattan_plots}). This suggests that GWAS focused on finding strongly associated alleles for traits controlled by a limited number of variants in the genome are likely robust to the impacts of continuous spatial structure. However, methods that analyze the combined effects of thousands or millions of weakly associated variants such as polygenic risk scores \citep{wray2007prediction,international2009common} are likely to be affected by subtle population structure. Indeed as recently identified in studies of genotype associations for human height in Europe \citep{Berg2018,Sohail2018}, PC regression GWAS in modern human populations do include residual signal of population structure in large-scale analyses of polygenic traits. In addition to error associated with varying patterns of linkage disequilibrium and allele frequency among populations, the confounding of environmental and genetic effects on phenotypes introduced by population structure is expected to lead to low predictive power when polygenic scores are generated for populations outside the original GWAS cohort, \llabel{ldcaveats}
%When attempting to make predictions across populations with different environmental exposures, polygenic risk scores affected by population structure can be expected to offer low predictive power, 
as was shown in a recent study finding lower polygenic score predictive power outside European populations \citep{Martin2019}.
